---
title: "optimal contrasts"
author: "Carina Miller"
date: "2025-08-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Optimal Contrasts
(see vignette binary)

The formula for the optimal contrasts is given by
$$
\begin{aligned}
c^{\textrm{opt}} \propto
S^{-1}\biggl(\mu^0_m - \frac{(\mu^0_m)^T S^{-1}1_K}{1_K^T S^{-1} 1_K}\biggr)
\end{aligned}
$$
where $\mu^0_m$ is the standardized mean response,
$K$ is the number doses,
and $1_K$ is an all-ones vector of length $K$ and $S$ is the
covariance matrix of the estimates at the doses
[see @pinheiro2014].\
(from Vignette binary)\

"For each candidate model, an optimal contrast $c^{\textrm{opt}}$ that maximizes the probability of rejecting the hypothesis of non-signal dose-response is derived assuming that the candidate model is correct" (aus: Improved inference... Diniz)
```{r, fig.width = 6.9, out.width = '100%'}
contr <- optContr(mods, doses = doses_treat, S = cov, placAdj = T)
contr # is no longer a contrast matrix (columns do not sum to 0) 
plotContr(contr, xlab = "Dose", ylab = "Contrast coefficients")
```