---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DoseFinding

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/DoseFinding)](https://CRAN.R-project.org/package=DoseFinding)
[![R-CMD-check](https://github.com/openpharma/DoseFinding/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openpharma/DoseFinding/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The DoseFinding package provides functions for the design and analysis of dose-finding experiments (for example pharmaceutical Phase II clinical trials). It provides functions for: multiple contrast tests, fitting non-linear dose-response models, a combination of testing and dose-response modelling and calculating optimal designs,
both for normal and general response variable.

## Installation

You can install the development version of DoseFinding from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bbnkmp/DoseFinding")
```

## Examples

### Performing multiple contrast tests

```{r example1, warning = FALSE, message=FALSE}
library(DoseFinding)
data(IBScovars)

## perform (model based) multiple contrast test
## define candidate dose-response shapes
models <- Mods(linear = NULL, emax = 0.2, quadratic = -0.17,
               doses = c(0, 1, 2, 3, 4))
## plot models
plot(models)
## perform multiple contrast test
MCTtest(dose, resp, IBScovars, models=models,
                addCovars = ~ gender)
```

### Fitting non-linear dose-response model
```{r example2}
## fit non-linear emax dose-response model
fitemax <- fitMod(dose, resp, data=IBScovars, model="emax",
                  bnds = c(0.01,5))
## display fitted dose-effect curve
plot(fitemax, CI=TRUE, plotData="meansCI")

```

### Optimal designs for dose estimation
```{r example3}
## Calculate optimal designs for target dose (TD) estimation
doses <- c(0, 10, 25, 50, 100, 150)
fmodels <- Mods(linear = NULL, emax = 25, exponential = 85,
                logistic = c(50, 10.8811),
                doses = doses, placEff=0, maxEff=0.4)
plot(fmodels, plotTD = TRUE, Delta = 0.2)
weights <- rep(1/4, 4)
optDesign(fmodels, weights, Delta=0.2, designCrit="TD")
```
