---
title: "Analysis template MCP-Mod for continuous data"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE)
```

## Background

Provide background for NVA example (see PMX MCPMod training slides)

```{r}
library(DoseFinding)
doses <- c(0, 12.5, 25, 50, 100)
## define candidate models (as used in NVA study)
mods <- Mods(emax = c(2.6, 12.5), 
             sigEmax = c(30.5, 3.5),
             quadratic = -0.00776,
             placEff = 1.25, maxEff = 0.15,
         doses = doses)
## graphical display of candidate shapes
plot(mods)
```

## Generate example data based on summary estimates

```{r}
data(glycobrom) ## NVA summary data
head(glycobrom)
## simulate full data based on summary estimates
set.seed(1)
sigma <- sqrt(0.015^2*60)
rand <- rep(MASS::mvrnorm(60, 0, sigma^2, empirical = TRUE), 5)
NVA <- data.frame(dose = rep(glycobrom$dose, each=60),
                  FEV1 = rep(glycobrom$fev1, each = 60) + rand)
head(NVA)
```

## Display optimal contrasts

```{r}
## calculate optimal contrasts
optC <- optContr(mods, w=1)
## display and plot
optC
plot(optC)
```

## Run MCP-part of analysis
```{r}
## run multiple contrast test
MMtest <- MCTtest(dose, FEV1, models=mods, data=NVA)
## basic results (edited)
MMtest
```

Here there is a clear dose-response trend.

Now repeat testing based on generalized MCP-Mod (in practice it may
often be based on an MMRM model, see below for code).

## Dose-response estimation

The approach here mixes bootstrapping and model averaging. 
Bootstrapping is performed by sampling from the multivariate normal distribution of the estimates originating from
the first stage model (here an ANOVA model, but could also come from an MMRM model or a logistic regression, ...).
For each bootstrapped data-set then a model selection is performed using AIC and inferences for this model saved 
(e.g. estimated dose-response curve).
Inference at the end is then based on these bootstrapped estimates, for example the estimate of the dose-response
curve is the average or median of the bootstrap estimates at each dose (which then might be from different models).
Credible intervals can be derived based on bootstrap quantiles. Inference for other quantities of interest can be
derived similarly. As different models contribute to the bootstrap resamples, the approach is more robust than
model selection (see also [https://doi.org/10.1002/sim.6991] for the result of extensive simulations on this topic).
Note that contrary to the MCP step, here all model parameters are estimated.

```{r}
## first produce summary estimates
fit <- lm(FEV1~factor(dose)-1, data=NVA)
mn <- coef(fit)
S <- vcov(fit)
dose <- c(0,12.5,25,50,100)
## create bootstrap replicates
nSim <- 1000
sims <- rmvnorm(nSim, mn, S)
## now actual model fitting and selection of best model for all bootstraps
bnds <- defBnds(max(dose)) ## boundaries assumed for model fitting (force convergence)
pred <- matrix(nrow=nSim, ncol=101) ## will contain dr-model predictions 
fit <- vector("list", 3)
for(i in 1:nSim){
  fit[[1]] <- fitMod(dose, sims[i,], model="emax", S=S,
                     type="general", bnds=bnds$emax)
  fit[[2]] <- fitMod(dose, sims[i,], model="sigEmax", S=S,
                     type="general", bnds=bnds$sigEmax)
  fit[[3]] <- fitMod(dose, sims[i,], model="quadratic", S=S,
                     type="general", bnds=bnds$quadratic)
  ind <- which.min(sapply(fit, gAIC)) ## choose best model according to AIC
  pred[i,] <- predict(fit[[ind]], doseSeq=0:100, predType="ls-means")
}
qq <- apply(pred, 2, function(x) quantile(x, c(0.025,0.25,0.5,0.75,0.975)))
matplot(0:100, t(qq), type="l", lty=c(3,2,1,2,3), col=1, lwd=2,
        xlab="Dose", ylab="Mean FEV1")
points(dose, tapply(NVA$FEV1, NVA$dose, mean), pch=19, cex=1.5)
```
