---
title: "Sample size template for MCP-Mod for normally distributed data"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE)
```

Provide useful power plots (maybe even going beyond what exists in the
package now, e.g. using ggplot2), that can easily be re-used for other
purposes.

# Power calculations

```{r}
library(DoseFinding)
doses <- c(0, 12.5, 25, 50, 100)
## define candidate models (as used in NVA study)
mods <- Mods(emax = c(2.6, 12.5), 
             sigEmax = c(30.5, 3.5),
             quadratic = -0.00776,
             placEff = 1.25, maxEff = 0.15,
         doses = doses)
## graphical display of candidate shapes
plot(mods)
```

```{r}
## optimal contrasts
contMat <- optContr(mods, w=1)
pows <- powN(upperN = 100, lowerN=10, step = 10,  
             contMat = contMat,sigma = 0.34,  
             altModels = mods, alpha = 0.05, 
             alRatio = rep(1, 5))
plot(pows)
## wrapper function sampSize performs sample-size calculations
sampSizeMCT(upperN=100, contMat = contMat, sigma = 
            0.34, altModels=mods, power = 0.9, 
alRatio = rep(1, 5), alpha = 0.05, sumFct = mean)
```

Here add something on power for mis-specified scenarios?


# Sample size based on metrics other than power

Refer to Chapter 12 in [@oquigley:2017] for example code on how to
evaluate precision in dose-response and dose estimate, using the
DoseFinding package.
