---
title: "Sample size calculations with MCP-Mod"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Sample size template for MCP-Mod for normally distributed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/settings.txt"}
```

<!-- TODO
- [ ] Where does the `sigma=0.34` come from? (Björn)
- [ ] Is the mis-specified exponential model 'bad enough'? (Björn)
-->

In this vignette we'll take a closer look at the design stage
and see how to perform power and sample size calculations with the DoseFinding package.

We'll consider the same example study
and the same candidate models as in the [basic usage vignette](2_analysis_normal.html).

```{r, setup}
library(DoseFinding)
library(ggplot2)
doses <- c(0, 12.5, 25, 50, 100)
guess <- list(emax = c(2.6, 12.5), sigEmax = c(30.5, 3.5), quadratic = -0.00776)
mods <- do.call(Mods, append(guess, list(placEff = 1.25, maxEff = 0.15, doses = doses)))
plot(mods)
```

## Power versus group sample size

In this section we'll look at how power varies with sample size.
Note that the effect size is fixed through `maxEff` in the candidate models.

First we calculate the matrix of optimal contrasts
(`w=1` denotes homoscedastic residuals with equal group sizes, see `?optContr`).

In `powN` we specify the sample sizes for which to calculate the power.
We also request five equally sized groups with `alRatio = rep(1, 5)`.
We fix the residual standard deviation with `sigma = 0.34`.

```{r, power_sample_size_1}
contMat <- optContr(mods, w=1)
pows <- powN(upperN = 100, lowerN = 10, step = 10, contMat = contMat,
             sigma = 0.34, altModels = mods, alpha = 0.05, alRatio = rep(1, 5))
plot(pows)
```

There also is a wrapper function that calculates the group sample sizes needed in order to attain a specific power.
The powers under each alternative model are combined with `sumFct`.
Here we look at the minimum power,
other potential choices are `mean` or `max`.

```{r, power_sample_size_2}
sampSizeMCT(upperN = 150, contMat = contMat, sigma = 0.34, altModels = mods,
            power = 0.9, alRatio = rep(1, 5), alpha = 0.05, sumFct = min)
```

## Power versus treatment effect

In this section we fix the group sample size at 60 and vary the treatment Effect `maxEff`.
Note how power decreases if we assume a higher residual standard deviation.

```{r, power_effect_size}
plot_power_vs_treatment_effect <- function(guess, doses, group_size, placEff, maxEffs,
                                           sigma_low, sigma_high, alpha) {
  mods_args_fixed <- append(guess, list(placEff = placEff, doses = doses))
  grid <- expand.grid(max_eff = maxEffs, sigma = c(sigma_low, sigma_high))
  grid$min_power <- grid$mean_power <- NA
  for (i in 1:nrow(grid)) {
    mods <- do.call(Mods, append(mods_args_fixed, list(maxEff = grid$max_eff[i])))
    p <- powMCT(optContr(mods, w = 1), alpha, mods, group_size, grid$sigma[i])
    grid$min_power[i] <- min(p)
    grid$mean_power[i] <- mean(p)
  }
  subt <- sprintf("placEff = %.3f, group size = %d, α = %.3f", placEff, group_size, alpha)
  gg <- ggplot(grid) + geom_line(aes(max_eff, min_power, linetype = "minimum")) +
    geom_line(aes(max_eff, mean_power, linetype = "mean")) +
    xlab("maximum treatment effect") + ylab("power") +
    facet_wrap(vars(sigma), labeller = label_both) +
    labs(title = "Power vs effect size for different residual standard deviations", subtitle = subt) +
    scale_linetype_discrete(name = "aggregation") +
    theme(legend.position = "bottom") +
    coord_cartesian(ylim = c(0, 1))
  return(gg)
}

plot_power_vs_treatment_effect(guess, doses, group_size = 60, placEff = 1.25,
                               maxEffs = seq(0.01, 0.3, length.out = 10),
                               sigma_low = 0.34, sigma_high = 0.68, alpha = 0.05)
```

## Power under mis-specification

One of the motivations behind MCP-Mod is robustness against mis-specified alternative models.
To illustrate this, let's include an exponential candidate model with only small responses for all doses but the last.

```{r, power_miss_1}
guess_miss <- list(exponential = guesst(25, 0.025, "exponential", Maxd = max(doses)))
mods_miss <- do.call(Mods, c(guess, guess_miss, list(placEff = 1.25, maxEff = 0.15, doses = doses)))
plot(mods_miss, superpose = TRUE)
```

Now we compare the power calculations with those based on the original candidate set.

```{r, power_miss_2}
plot_power_misspec <- function(guess, guess_miss, placEff, maxEff, doses,
                               upperN, lowerN, step, sigma, alpha) {
  mods_extra_par <- list(placEff = placEff, maxEff = maxEff, doses = doses)
  pown_extra_par <- list(upperN = upperN, lowerN = lowerN, step = step,
                         sigma = sigma, alpha = alpha, alRatio = rep(1, length(doses)))
  mods_miss <- do.call(Mods, c(guess, guess_miss, mods_extra_par))
  mods_ok <- do.call(Mods, c(guess, mods_extra_par))
  cm_miss <- optContr(mods_miss, w = 1)
  cm_ok <- optContr(mods_ok, w = 1)
  p_miss <- do.call(powN, c(pown_extra_par, list(contMat = cm_miss, altModels = mods_miss)))
  p_ok <- do.call(powN, c(pown_extra_par, list(contMat = cm_ok, altModels = mods_ok)))
  pwr <- rbind(data.frame(n = as.numeric(rownames(p_ok)), p_ok[, c("min", "mean")], miss = FALSE),
               data.frame(n = as.numeric(rownames(p_miss)), p_miss[, c("min", "mean")], miss = TRUE))

  gg <- ggplot(pwr, aes(group = miss, color = miss)) +
    geom_line(aes(n, min, linetype = "minimum")) +
    geom_line(aes(n, mean, linetype = "mean")) +
    scale_color_discrete(name = "miss-specified") +
    scale_linetype_discrete(name = "aggregation") +
    labs(title = "Mean and minimum power under mis-specification") +
    xlab("group size") + ylab("power") +
    coord_cartesian(ylim = c(0, 1))
  return(gg)
}

plot_power_misspec(guess, guess_miss, placEff = 1.25, maxEff = 0.15, doses = doses,
                   upperN = 100, lowerN = 10, step = 10, sigma = 0.34, alpha = 0.05)

sampSizeMCT(upperN = 150, contMat = optContr(mods_miss, w = 1), sigma = 0.34, altModels = mods_miss,
            power = 0.9, alRatio = rep(1, 5), alpha = 0.05, sumFct = min)
```

Note how the mean power drops only slightly.
The required group size for 90% power only increases by 7.

## Sample size based on metrics other than power

Refer to Chapter 12 in @oquigley2017 for example code on how to evaluate precision in dose-response and dose estimation.

## References
