---
title: "Multiple Regimen MCP-Mod"
output: rmarkdown::html_vignette
bibliography: refs.bib
link-citations: yes
cls: american-statistical-association.csl
vignette: >
  %\VignetteIndexEntry{Analysis template: MCP-Mod with multiple regimen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, child="children/settings.txt"}
```

## Background

Often more than one regimen is studied in dose-finding studies. If there
are enough doses within each regimen, one may still utilize MCP-Mod.
But specific assumptions are needed, and it depends on the situation,
whether or not these are appropriate (and thus usage of MCP-Mod).

The first idea is to bring the doses for each regimen on a common
scale (total dose per time unit). For example if once daily (o.d.)
dosing and twice daily (b.i.d.) dosing are used in a study, one might
utilize the total daily dose.

It is usually not appropriate to then perform MCP-Mod on the total daily
dose (ignoring from which regimen the doses originate): The study
investigated more than one regimen, so assessing the difference
between regimen (for example for the same total daily dose) is of
interest. This would not be possible with a modelling approach
that ignores the regimen.

The most general approach would be to perform MCP-Mod separately by
regimen, and for example to adjust p-values originating from the
MCP-part using a Bonferroni-Holm correction. This approach assumes that the
regimen don't share any similarity. This approach is often not
feasible, due to the double-blind nature of trials. In the example of
o.d. and b.i.d., all patients would receive two administrations per
day (patients in the o.d. group receive one placebo per day), so that
there is no real o.d. group, and no placebo o.d. group.  The placebo
group is then assumed common to both the o.d. and b.i.d. dose-response
curve. For the MCP-step contrasts for both o.d. and b.i.d. are taken
with respect to the same placebo group and in the modelling step one
would then assume the intercept to be the same across regimen, but all
other parameters separate.

One could also assume further parameters to be common across regimen
(for example the Emax parameter for the Emax model), but in the
following example no such assumption is made.

The motivation for the simulated data below is taken from a recently
completed dose-finding study, where the dose-response of the drug
Licogliflozin was assessed for the o.d. and b.i.d. regimen [@bays2020],
see also the [corresponding page at clinicaltrials.gov](https://clinicaltrials.gov/ct2/show/results/NCT03100058).

Note that this study used MCP-Mod, but the analysis presented here has
been modified and simplified (e.g. in terms of candidate models and
dose-response modelling strategy).

$\mu=(\mu_{\mathrm{placebo}}, \mu_{\mathrm{od}}, \mu_{\mathrm{bid}})$

```{r, data}
library(DoseFinding)
library(ggplot2)
# collect estimates and dosage information in one place
example_estimates <- function() {
  # ANOVA mean estimates and ci bounds extracted from fig. 3 of Bays (2020),
  # because the clinicaltrials.gov page seems to contain values from a parametric model fit
  mn <- c(-0.55, -1.78, -1.95, -3.29, -4.43, -1.14, -2.74, -4.03, -4.47)
  lb <- c(-1.56, -3.15, -3.36, -4.85, -5.40, -2.49, -4.10, -5.50, -5.50)
  ub <- c( 0.40, -0.30, -0.54, -1.76, -3.48, 0.24, -1.38, -2.65, -3.44)
  se <- (ub - lb)/(2*qnorm(0.975)) # approximate standard error
  return(list(mu_hat = mn,
              daily_dose = c(0, 2.5, 10, 50, 150, 5, 10, 50, 100),
              S_hat = diag(se^2),
              index = list(placebo = 1, od = 2:5, bid = 6:9)))
}

tidy_estimates <- function(est) {
  se <- sqrt(diag(est$S_hat))
  tidy_df <- data.frame(daily_dose = est$daily_dose,
                        mu_hat = est$mu_hat,
                        ub = est$mu_hat + qnorm(0.975) * se,
                        lb = est$mu_hat - qnorm(0.975) * se)
  tidy_df <- rbind(tidy_df[1, ], tidy_df) # duplicate placebo
  tidy_df$regimen <- c("od", "bid", rep("od", length(est$index$od)), rep("bid", length(est$index$bid)))
  return(tidy_df)
}

plot_estimates <- function(est) {
  df <- tidy_estimates(est)
  ggplot(df, aes(daily_dose, mu_hat)) + geom_point() +
    geom_errorbar(aes(ymin = lb, ymax = ub)) +
    facet_wrap(vars(regimen), labeller = label_both) +
    xlab("daily dose") + ylab("percent body weight cange") +
    labs(title = "ANOVA estimates with 95% confindence intervals")
}

est <- example_estimates()
plot_estimates(est)
```


## Candidate models

Assume modelling is performed on the total daily dose scale, and that
the same candidate models are utilized for the two regimen.

```{r, candidate_models}
mods <- list(
  od = Mods(emax = c(5, 50),
            sigEmax = rbind(c(75,3.5), c(25, 0.7)),
            maxEff = -1,
            doses=c(0, est$od$daily_dose)),
  bid = Mods(emax = c(5, 50),
             sigEmax = rbind(c(75,3.5), c(25, 0.7)),
             maxEff = -1,
             doses=c(0, est$bid$daily_dose)))

plot(mods$od, superpose = TRUE)
plot(mods$bid, superpose = TRUE)
```

## Multiple contrast test

The matrix of contrasts is built from two separate matrices,
one for each regimen.

```{r, contrasts}
calculate_contrasts <- function(est, mods) {
  S_hat <- est$S_hat
  i <- est$index
  cm_od <- optContr(mods$od, S=S_hat[c(i$placebo, i$od), c(i$placebo, i$od)])$contMat
  cm_bid <- optContr(mods$bid, S=S_hat[c(i$placebo, i$bid), c(i$placebo, i$bid)])$contMat
  colnames(cm_od) <- paste0("od_", colnames(cm_od))
  rownames(cm_od)[-1] <- paste0("od_", rownames(cm_od)[-1])
  colnames(cm_bid) <- paste0("bid_", colnames(cm_bid))
  rownames(cm_bid)[-1] <- paste0("bid_", rownames(cm_bid)[-1])
  # now build a block matrix (contrasts in columns) like this:
  # [placebo od | placebo bid ]
  # [doses od   | 0           ]
  # [0          | doses bid   ]
  cm_full <- rbind(
    "0"=c(cm_od[1,],                                cm_bid[1,]                              ),
    cbind(cm_od[-1,],                               matrix(0, nrow(cm_od) - 1, ncol(cm_bid))),
    cbind(matrix(0, nrow(cm_bid) - 1, ncol(cm_od)), cm_bid[-1, ]                            ))
  return(cm_full)
}

cont_mat <- calculate_contrasts(est, mods)
print(round(cont_mat, 2))
```

We also need to calculate the test statistics by hand.

```{r, test}
mct_test <- function(cont_mat, est) {
  cont_cov <- t(cont_mat) %*% est$S_hat %*% cont_mat
  t_stat <- drop(est$mu_hat %*% cont_mat) / sqrt(diag(cont_cov))
  # FIXME: calling non-exported function
  p <- DoseFinding:::pValues(cont_mat, cov2cor(cont_cov), Inf, t_stat, "one.sided")
  ord <- rev(order(t_stat))
  return(data.frame(tStat = t_stat[ord], pVals = p[ord]))
}
mct_test(cont_mat, est)
```

## Dose-response modelling

First auxiliary code used for model fitting, based on the functions in
the MCP-Mod package.

```{r, estimation, warning=FALSE}
# evaluate model for od/bid with common e0, but separate remaining parameters
# arguments:
# - model: as a string like "emax",
# - i_par: list of vectors named "placebo", "od", "bid", used for indexing `par`
# - par: numeric, model parameter structured as c(e0, pars_od, pars_bid)
# returns: response at placebo, dose_od, dose_bid (in this order)
eval_model_shared_e0 <- function(model, dose_od, dose_bid, par, i_par) {
  resp_placebo <- par[1] # e0
  resp_od <- do.call(model, append(list(dose_od, par[1]), as.list(par[i_par$od])))
  resp_bid <- do.call(model, append(list(dose_bid, par[1]), as.list(par[i_par$bid])))
  return(c(resp_placebo, resp_od, resp_bid))
}

# fits 'model' to mu_hat with GLS (usint S_hat_inv as weight matrix), using a common e0 for od and bid regimens.
# i_reg:  list of vectors named "placebo", "od", "bid", used for indexing `dose`
# i_par: passed through to `eval_model_shared_e0`
# dose: numeric with doses for placebo, od, bid
# lower, upper, start: control parameters fro `nlminb`
fit_model_shared_e0 <- function(model, dose, mu_hat, S_hat_inv, lower, upper, start, i_reg, i_par) {
  opt_fun <- function(par) { # make use of lexical scope
    resp <- eval_model_shared_e0(model, dose[i_reg$od], dose[i_reg$bid], par, i_par)
    delta <- resp - mu_hat
    return(drop(t(delta) %*% S_hat_inv %*% delta))
  }
  fit <- nlminb(start, opt_fun, lower = lower, upper = upper)
  return(fit)
}

# find sensible starting values for `fit_model_shared_e0` by fitting separate models,
# index:  list of vectors named "placebo", "od", "bid", used for indexing `dose`
# bounds: passed through to `fitMod`
calc_start_values <- function(model, full_mu, full_S, dose, index, bounds) {
  separate_coefs <- sapply(c("od", "bid"), function(regimen) {
    which <- c(index$placebo, index[[regimen]])
    coef(fitMod(dose[which], full_mu[which], S = full_S[which, which],
                type = "general",  model = model, bnds = bounds))[-1] # drop e0 estimate
  })
  # remove names to prevent error in do.call() in eval_model_shared_e0;
  # od, bid coefs are in 1st / second column
  start <- c(full_mu[1], as.numeric(separate_coefs), use.names=FALSE)
  return(start)
}

# predict population response in each regimen for dose_seq_*
# note: both dose_seq_* vectors should contain a 0 if response at placebo is of interest
one_bootstrap_sample <- function(est, dose_seq_od, dose_seq_bid) {
  mu_new <- drop(rmvnorm(1, est$mu_hat, est$S_hat))
  mod_info <- list(list(name = "emax", bounds = rbind(c(0.15, 225)),
                        i_par = list(od = 2:3, bid = 4:5), n_par_gaic = 5),
                   list(name = "sigEmax", bounds = rbind(c(0.15, 225), c(0.5, 5)),
                        i_par = list(od = 2:4, bid = 5:7), n_par_gaic = 7))
  fit <- lapply(mod_info, function(m) {
    start <- calc_start_values(m$name, mu_new, est$S_hat, est$daily_dose, est$index, m$bounds)
    low <- c(-Inf, -Inf, m$bounds[,1]) # no bounds on e0, ed50
    up <- c(Inf, Inf, m$bounds[,2])
    fit_model_shared_e0(m$name, est$daily_dose, mu_new, solve(est$S_hat), lower = low,  upper = up,
                        start = start, i_reg = est$index, i_par = m$i_par)
  })
  gaics <- sapply(fit, `[[`, "objective") + 2 * sapply(mod_info, `[[`, "n_par_gaic")
  sel <- which.min(gaics)
  mod <- mod_info[[sel]]
  # drop the placebo element
  pred <- eval_model_shared_e0(mod$name, dose_seq_od, dose_seq_bid, fit[[sel]]$par, mod$i_par)[-1]
  return(pred)
}

summarize_bootstrap_samples <- function(samples, probs = c(0.025, 0.25, 0.75, 0.975)) {
  stopifnot(length(probs) == 4)
  med <- apply(samples, 1, median)
  quants <- apply(samples, 1, quantile, probs = probs, na.rm = TRUE) # FIXME: there shouldn't be any NAs here
  bs_df <- as.data.frame(cbind(med, t(quants)))
  names(bs_df) <- c("median", "low_out", "low_in", "high_in", "high_out")
  return(bs_df)
}

dose_seq_od <- seq(0, 150, length.out = 21) # do include placebo!
dose_seq_bid <- seq(0, 100, length.out = 21)
set.seed(1)
reps <- replicate(25, one_bootstrap_sample(est, dose_seq_od, dose_seq_bid))
bs_sum <- summarize_bootstrap_samples(reps)
bs_sum$daily_dose <- c(dose_seq_od, dose_seq_bid)
bs_sum$regimen <- c(rep("od", length(dose_seq_od)), rep("bid", length(dose_seq_bid)))

ggplot(bs_sum) + geom_ribbon(aes(daily_dose, ymin=low_out, ymax=high_out), alpha = 0.2) +
  geom_ribbon(aes(daily_dose, ymin=low_in, ymax=high_in), alpha = 0.2) +
  geom_line(aes(daily_dose, median)) +
  geom_point(aes(daily_dose, mu_hat), tidy_estimates(est)) +
  facet_wrap(vars(regimen), labeller = label_both) +
  labs(title = "Bootstrap estimates for population response") +
  xlab("daily dose") + ylab("percent body weigh change") +
  coord_cartesian(ylim = c(-6, 0))
```
