---
title: "Sample size calculations with MCP-Mod"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Sample size template for MCP-Mod for normally distributed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/settings.txt"}
```

In this vignette we'll take a closer look at the design stage and see
how to perform power and sample size calculations for MCP-Mod with the
DoseFinding package.

We'll consider the same example study
and the same candidate models as in the [basic usage vignette](analysis_normal.html).

```{r, setup, fig.asp = 1, out.width = "50%", fig.width = 5}
library(DoseFinding)
library(ggplot2)
doses <- c(0, 12.5, 25, 50, 100)
guess <- list(emax = c(2.6, 12.5), sigEmax = c(30.5, 3.5), quadratic = -0.00776)
mods <- do.call(Mods, append(guess, list(placEff = 1.25, maxEff = 0.15, doses = doses)))
plot(mods)
```

## Power versus group sample size

In this section we'll look at how power varies with sample size.  Note
that the maximum effect size over the dose-range is fixed through
`maxEff` in the candidate models.

First we calculate the matrix of optimal contrasts
(`w=1` denotes homoscedastic residuals with equal group sizes, see `?optContr`).

In `powN` we specify the sample sizes for which to calculate the power.
We also request five equally sized groups with `alRatio = rep(1, 5)`.
We fix the residual standard deviation with `sigma = 0.34`.

```{r, power_sample_size_1}
contMat <- optContr(mods, w=1)
pows <- powN(upperN = 100, lowerN = 10, step = 10, contMat = contMat,
             sigma = 0.34, altModels = mods, alpha = 0.05, alRatio = rep(1, 5))
plot(pows)
```
This shows the power values of the maximum contrast test assuming each
of the different candidate models to be true. The minimum, mean and
maximum power over the candidate models are also included in the plot.

There also is a wrapper function that calculates the group sample
sizes needed in order to attain a specific power.  The powers under
each alternative model are combined with `sumFct`.  Here we look at
the minimum power, other potential choices are `mean` or `max`.

```{r, power_sample_size_2}
sampSizeMCT(upperN = 150, contMat = contMat, sigma = 0.34, altModels = mods,
            power = 0.9, alRatio = rep(1, 5), alpha = 0.05, sumFct = min)
```

## Power versus treatment effect

In this section we fix the group sample size at 90 and vary the treatment Effect `maxEff`.
Note how power decreases if we assume a higher residual standard deviation.

```{r, power_effect_size}
plot_power_vs_treatment_effect <- function(guess, doses, group_size, placEff, maxEffs,
                                           sigma_low, sigma_mid, sigma_high, alpha) {
  mods_args_fixed <- append(guess, list(placEff = placEff, doses = doses))
  grd <- expand.grid(max_eff = maxEffs, sigma = c(sigma_low, sigma_mid, sigma_high))
  min_power <- mean_power <- NA
  for (i in 1:nrow(grd)) {
    mods <- do.call(Mods, append(mods_args_fixed, list(maxEff = grd$max_eff[i])))
    p <- powMCT(optContr(mods, w = 1), alpha, mods, group_size, grd$sigma[i])
    min_power[i] <- min(p)
    mean_power[i] <- mean(p)
  }
  grd$sigma <- factor(grd$sigma)
  pdat <- cbind(grd, power = c(min_power, mean_power),
                sumFct = rep(factor(1:2, labels = c("min", "mean")), each = nrow(grd)))
  subt <- sprintf("placEff = %.3f, group size = %d, α = %.3f", placEff, group_size, alpha)
  gg <- ggplot(pdat) + geom_line(aes(max_eff, power, lty = sigma)) +
    facet_wrap(~sumFct, labeller = label_both)+
    xlab("maximum treatment effect") + ylab("power") +
    labs(title = "Minimum power vs effect size for different residual standard deviations", subtitle = subt) +
    theme(legend.position = "bottom") +
    scale_y_continuous(limits = c(0,1), breaks = seq(0,1,by=.1))
  return(gg)
}

plot_power_vs_treatment_effect(guess, doses, group_size = 90, placEff = 1.25,
                               maxEffs = seq(0.01, 0.3, length.out = 15),
                               sigma_low = 0.3, sigma_mid = 0.34, sigma_high = 0.4, alpha = 0.05)
```

## Power under mis-specification

One of the motivations behind MCP-Mod is robustness against
mis-specified alternative models.  To illustrate this, let's assume an
exponential model shape is the true model, which only has small
responses for all doses but the last (here assuming 20% of the overall
treatment effect is achieved at the 50μg dose). All other candidate
shapes that almost the full effect is achieved for the 50μg dose, so
this shape is quite different from all other shapes included in the
candidate set.

```{r, power_miss_1}
guess_miss <- list(exponential = guesst(50, 0.2, "exponential", Maxd = max(doses)))
mods_miss <- do.call(Mods, c(guess, guess_miss, list(placEff = 1.25, maxEff = 0.15, doses = doses)))
plot(mods_miss, superpose = TRUE)
```

Now we compare the power calculation under the exponential model with
those based on the original candidate set, in both cases only the
contrasts from the original candidate set are used.

```{r, power_miss_2}
plot_power_misspec <- function(guess, guess_miss, placEff, maxEff, doses,
                               upperN, lowerN, step, sigma, alpha) {
  mods_extra_par <- list(placEff = placEff, maxEff = maxEff, doses = doses)
  pown_extra_par <- list(upperN = upperN, lowerN = lowerN, step = step,
                         sigma = sigma, alpha = alpha, alRatio = rep(1, length(doses)))
  mods_miss <- do.call(Mods, c(guess_miss, mods_extra_par))
  mods_ok <- do.call(Mods, c(guess, mods_extra_par))
  cm_ok <- optContr(mods_ok, w = 1)
  p_miss <- do.call(powN, c(pown_extra_par, list(contMat = cm_ok, altModels = mods_miss)))
  p_ok <- do.call(powN, c(pown_extra_par, list(contMat = cm_ok, altModels = mods_ok)))
  pwr <- rbind(data.frame(n = as.numeric(rownames(p_ok)), p_ok[, c("min", "mean")], miss = FALSE),
               data.frame(n = as.numeric(rownames(p_miss)), p_miss[, c("min", "mean")], miss = TRUE))

  gg <- ggplot(pwr, aes(group = miss, color = miss)) +
    geom_line(aes(n, min, linetype = "minimum")) +
    geom_line(aes(n, mean, linetype = "mean")) +
    scale_color_discrete(name = "miss-specified") +
    scale_linetype_discrete(name = "aggregation") +
    labs(title = "Mean and minimum power under mis-specification") +
    xlab("group size") + ylab("power") +
    scale_y_continuous(limits = c(0,1), breaks = seq(0,1,by=.1))
  return(gg)
}

plot_power_misspec(guess, guess_miss, placEff = 1.25, maxEff = 0.15, doses = doses,
                   upperN = 100, lowerN = 10, step = 10, sigma = 0.34, alpha = 0.05)
```

It can be seen that, as expected, the power drops as the exponential
model shape is quite different from all candidate shapes included, but
the power loss is only in the range of 10-15%.

## Sample size based on metrics other than power

The main purpose of a dose-finding study is selection of a dose to
take forward into Phase 3, so why is sample size calculation based on
the MCP-part of MCP-Mod even relevant?  The multiple contrast test is
the minimum requirement, before even considering dose selection. It
turns out, however, that dose selection is a much harder problem than
establishing a dose-response effect versus placebo based on the
MCP-part: The sample size required for adequate accuracy in estimation
of a target dose (e.g. the smallest dose achieving a relevant
improvement over placebo) is usually several-fold higher than the
sample size needed to have adequate power for the MCP-part.  This
should not come as a surprise as dose-selection is primarily a
comparison among the investigational doses, while the MCP-part
establishes an effect versus placebo.  Chapter 12 in @oquigley2017
illustrates this with simulations, based on the `planMod` function
(see `?planMod` for example usage). This consideration should be kept
in mind for sample size calculation. For example the `planMod`
function could be used to evaluate the dose estimation performance for
a given sample size.

On the other hand the efficacy dose-response curve is not the only
consideration in dose-selection: Results for other efficacy/biomarker
endpoints, but also the results for tolerability or safety markers,
will contribute to that decision.

## References
